Using gringo, clingo and iclingo \\\small{Rev.\svnInfoRevision}

### 2008-11-08 doc/documentation/restrictions.tex
polished restrictedness notions
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@590 3e02f596-3131-0410-8352-e04eeaec37cc

In view of function symbols with non-zero arity,
we may be confronted with logic programs over an infinite Herbrand base.
In order to maintain decidability of reasoning tasks,
it is thus important to identify language fragments for which
finite equivalent ground programs are guaranteed to exist.
Level-restricted logic programs constitute such a fragment,
where finiteness is manifested in the requirement that any variable in a rule
must be bound to a finite set of ground terms via a predicate not
subject to positive recursion through the rule.
The notion of level-restrictedness is complemented by stratification,
which by disallowing negative recursion among predicates describes a
class of logic programs having unique answer sets.
The formal definition of level-restricted programs
that can be grounded by \gringo, \clingo, and \iclingo\
is provided in Section~\ref{subsec:lambda}.
Section~\ref{subsec:strat} introduces stratified programs and the related
concept of domain predicates, which can serve particular purposes during grounding.
Both sections may be skimmed or even skipped upon the first reading,
and rather be looked up later on to find out details.

@@

The main task of a grounder is to substitute the variables in a logic program~$\Pi$
by terms such that the result is a finite equivalent ground program~$\alt{\Pi}$.%
\footnote{$\Pi$ and~$\alt{\Pi}$ are equivalent if they have the same answer sets.}
Of course, a necessary condition for this is that~$\Pi$ possesses
only finitely many finite answer sets.
Unfortunately, such properties are undecidable in general~\cite{daeigovo01a}.
Instead of semantic properties, grounders do thus impose rather simple syntactic
conditions to guarantee the existence of finite equivalent ground programs
(which are then also computed).
For instance, the \dlv\ system~\cite{dlv03a} requires programs to be \emph{safe} for
establishing finiteness in the absence of functions with non-zero arity and under
limited arithmetic.
Grounder \lparse~\cite{lparseManual} deals with \emph{$\omega$-restricted}
programs~\cite{syrjanen01a}, which are more restrictive than safe programs
but guarantee finiteness also in the presence of functions with non-zero arity.
Finally, \gringo~\cite{gescth07a} requires programs to be \emph{$\lambda$-restricted},
a property more general than $\omega$-restrictedness but likewise applicable
to programs over the same languages.
We below reproduce the concept of $\lambda$-restrictedness,
called level-restrictedness here.

The idea underlying level-restrictedness is that the structure of a logic program~$\Pi$
must be such that, for each rule $r\in\Pi$,
the set of potentially applicable ground instances of~$r$ is known in advance.
This is the case if, for each variable~$V\in\vars{r}$,
we find an atom~$A$ in the body of~$r$ with
$V\in\vars{A}$, where $\vars{A}$ is that set of variables occurring in~$A$,
such that the potentially derivable ground instances of~$A$ are limited (see below).
In fact, with such an atom~$A$ at hand,
the set of ground terms to which~$V$ needs to be instantiated is known a priori,
and no further ground terms need to be considered.
The sketched approach is justified by the $\subseteq$-minimality of answer sets
in the sense of Section~\ref{subsec:semantics},
as it allows us to restrict the attention to rules whose bodies are derivable w.r.t.\
some answer set.
Beyond outputting only ``relevant'' rules,
grounders can apply answer-set preserving simplifications,
some of which are discussed in Section~\ref{subsec:strat}.

@@

After dealing with stratified programs, we consider subprograms.
For a given logic program~$\Pi$, some~$\pi\subseteq\Pi$ is a 
\emph{stratified subprogram} of~$\Pi$ if
$\pi$ is stratified and if
no predicate~$p/i$ occurring in~$\pi$ belongs to the head of any rule in~$\Pi\setminus\pi$.
As a matter of fact,
a stratified normal subprogram~$\pi$ of~$\Pi$ has a unique answer set~$Y$ such that
$Y\subseteq X$ for any answer set~$X$ of~$\Pi$~\cite{liftur94a}.
A stratified subprogram~$\pi$ of~$\Pi$ is \emph{maximal} if every stratified
subprogram of~$\Pi$ is contained in~$\pi$.
Note that every logic program has a maximal stratified subprogram,
which is also easy to determine.
In fact, \gringo\ identifies the maximal stratified subprogram~$\pi$
of a logic program~$\Pi$,
and it fully evaluates the predicates~$p/i$ that occur in~$\pi$,
which we call \emph{domain predicates}.
In Section~\ref{subsec:gringo:condition},
we will see that domain predicates can serve particular purposes during grounding.

@@

One can check that the program consisting of Line~1--5
in Example~\ref{ex:strat:prog} is level-restricted, but not stratified.
Conversely, program
\begin{lstlisting}[numbers=none]
nat(0).  nat(s(X)) :- nat(X).
\end{lstlisting}
is stratified (take $\xi=\{\pred{nat}/1 \mapsto 0\}$),
but not level-restricted
($\lambda(\pred{nat}/1)>\lambda(\pred{nat}/1)$
 needed because of the second rule is impossible).
In fact, the program has a unique but infinite answer set, thus,
the program cannot be translated to a finite equivalent ground program.
For this reason, \gringo\ does not accept the given program.
Rather, in the first place, a program must be level-restricted in order
to ground it with \gringo.
If level-restrictedness is granted,
the maximal stratified subprogram is exploited in a second stage
to optimize grounding and to make use of domain predicates.

>>>>

In view of function symbols with non-zero arity,
we may be confronted with logic programs over an infinite Herbrand base.
In order to maintain decidability of reasoning tasks,
it is thus important to identify language fragments for which
finite equivalent ground programs are guaranteed to exist.
Level-restricted logic programs constitute such a fragment,
where finiteness is manifested in the requirement that any variable in a rule
must be bound to a finite set of ground terms via a predicate not
subject to positive recursion through that rule.
The notion of level-restrictedness is complemented by stratification,
which by disallowing negative recursion among predicates describes a
class of logic programs having unique answer sets.
The formal definition of level-restricted programs,
which can be grounded by \gringo\ (stand-alone or embedded in \clingo\ and \iclingo),
is provided in Section~\ref{subsec:lambda}.
Section~\ref{subsec:strat} introduces stratified programs and the related
concept of domain predicates, which can serve particular purposes during grounding.
Both sections may be skimmed or even skipped upon the first reading,
and rather be looked up later on to find out details.

@@

The main task of a grounder is to substitute the variables in a logic program~$\Pi$
by terms such that the result is a finite equivalent ground program~$\alt{\Pi}$.%
\footnote{$\Pi$ and~$\alt{\Pi}$ are equivalent if they have the same answer sets.}
Of course, a necessary condition for this is that~$\Pi$ possesses
only finitely many finite answer sets.
Unfortunately, such a property is undecidable in general~\cite{daeigovo01a}.
Instead of semantic properties, grounders do thus impose rather simple syntactic
conditions to guarantee the existence of finite equivalent ground programs
(which are then also computed).
For instance, the \dlv\ system~\cite{dlv03a} requires programs to be \emph{safe} for
establishing finiteness in the absence of functions with non-zero arity and under
limited arithmetic.
Grounder \lparse~\cite{lparseManual} deals with \emph{$\omega$-restricted}
programs~\cite{syrjanen01a}, which are more restrictive than safe programs
but guarantee finiteness also in the presence of functions with non-zero arity.
Finally, \gringo~\cite{gescth07a} requires programs to be \emph{$\lambda$-restricted},
a property more general than $\omega$-restrictedness but likewise applicable
to programs over the same languages.
We below reproduce the concept of $\lambda$-restrictedness,
called level-restrictedness here.

The idea underlying level-restrictedness is that the structure of a logic program~$\Pi$
must be such that, for each rule $r\in\Pi$,
the set of potentially applicable ground instances of~$r$ is known in advance.
This is the case if, for each variable~$V\in\vars{r}$,
we find an atom~$A$ in the body of~$r$ with
$V\in\vars{A}$ (where $\vars{A}$ is that set of variables occurring in~$A$)
such that the potentially derivable ground instances of~$A$ are limited (see below).
In fact, with such an atom~$A$ at hand,
the set of ground terms to which~$V$ needs to be instantiated is known a priori,
and no further ground terms need to be considered.
The sketched approach is justified by the $\subseteq$-minimality of answer sets
in the sense of Section~\ref{subsec:semantics},
as it allows us to restrict the attention to rules whose bodies are derivable w.r.t.\
some answer set.
Beyond outputting only ``relevant rules,''
grounders can apply answer-set preserving simplifications,
some of which are discussed in Section~\ref{subsec:strat}.

@@

After dealing with stratified programs, we consider subprograms.
For a given logic program~$\Pi$, some~$\pi\subseteq\Pi$ is a 
\emph{stratified subprogram} of~$\Pi$ if
$\pi$ is stratified and if
no predicate~$p/i$ occurring in~$\pi$ belongs to the head of any rule in~$\Pi\setminus\pi$.
As a matter of fact,
a stratified normal subprogram~$\pi$ of~$\Pi$ has a unique answer set~$Y$ such that
$Y\subseteq X$ for any answer set~$X$ of~$\Pi$~\cite{liftur94a}.
A stratified subprogram~$\pi$ of~$\Pi$ is \emph{maximal} if every stratified
subprogram of~$\Pi$ is contained in~$\pi$.
Note that every logic program has a maximal stratified subprogram,
which is also easy to determine.
In fact, \gringo\ identifies the maximal stratified subprogram~$\pi$
of a logic program~$\Pi$,
and it fully evaluates the predicates~$p/i$ not belonging to the head of any rule in~$\Pi\setminus\pi$,
which we call \emph{domain predicates}.
In Section~\ref{subsec:gringo:condition},
we will see that domain predicates can serve particular purposes during grounding.

@@

One can check that the program consisting of Line~1--5
in Example~\ref{ex:strat:prog} is level-restricted, but not stratified.
Conversely, program
\begin{lstlisting}[numbers=none]
nat(0).  nat(s(X)) :- nat(X).
\end{lstlisting}
is stratified (take $\xi=\{\pred{nat}/1 \mapsto 0\}$),
but not level-restricted
($\lambda(\pred{nat}/1)>\lambda(\pred{nat}/1)$
 needed because of the second rule is impossible).
In fact, the program has a unique but infinite answer set.
Thus,
the program cannot be translated to a finite equivalent ground program,
which is why \gringo\ does not accept it.
Rather, in the first place, a program must be level-restricted in order
to ground it with \gringo.
If level-restrictedness is granted,
the maximal stratified subprogram is exploited in a second stage
to optimize grounding and to make use of domain predicates.
